---
title: "datascience_2_midterm"
output:
  pdf_document: default
  html_document: default
date: "2025-03-27"
---
```{r loading libraries}
library(dplyr)
library(tidyverse)
library(survival)
library(ggsurvfit)
library(gtsummary)
library(glue)
library(labelled)
library(rpart.plot)
library(rpart)
library(party)
library(partykit)
library(caret)
```

```{r loading data sets}
load("dat1.RData")
load("dat2.RData")
```

```{r adding labels}
#mutating factor variables and creating labels for baseline characteristic table
table_data = dat1 |>
  mutate(gender = factor(gender, levels = c(0,1), labels = c("Female","Male")),
         race = factor(race, levels = c(1,2,3,4), labels = c("White","Asian","Black","Hispanic")),
         smoking = factor(smoking, levels = c(0,1,2), labels = c("Never smoked", "Former smoker", "Current smoker")),
         diabetes = factor(diabetes, levels = c(0,1), labels = c("No","Yes")),
         hypertension = factor(hypertension, levels = c(0,1), labels = c("No", "Yes")))

#creating table labels
var_label(table_data) = list(
  age = "Age (in years)",
  gender = "Gender",
  race = "Race/ethnicity",
  smoking = "Smoking status",
  height = "Height (in centimeters)",
  weight = "Weight (in kilograms)",
  bmi = "BMI (body mass index)",
  diabetes = "Diabetes",
  hypertension = "Hypertension",
  SBP = "Systolic Blood Pressure (mmHg)",
  LDL = "LDL Cholesterol (mg/dL)",
  time = "Time since vaccination (in days)"
)
```

```{r baseline characteristic table}
#outputting table
table = table_data |>
  select(age, gender, race, smoking, height, weight, bmi, diabetes, hypertension, SBP, LDL, time) |>
  tbl_summary(
    missing_text = "(Missing)",
    statistic = list(all_continuous() ~ "{median} ± {sd}")
  ) 

print(table)
```

```{r data visualization of association between predictors and log antibody levels}
##Distribution of Log Antibody Levels by Race/Ethnicity
age_antibody <- ggplot(table_data, aes(x = age, y = log_antibody)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(x = "Age", y = "Average Log Antibody Level", 
       title = "Average Log Antibody Levels by Age") +
  theme_minimal()

age_antibody

#Average Log Antibody Levels by Gender"
gender_antibody <- ggplot(table_data, aes(x = factor(gender), y = log_antibody)) +
  geom_boxplot() +
  labs(x = "Gender", y = "Log Antibody Level") +
  theme_minimal() 

gender_antibody

#Distribution of Log Antibody Levels by Race/Ethnicity
race_antibody <- ggplot(table_data, aes(x = log_antibody, fill = factor(race))) +
  geom_histogram(binwidth = 0.1, position = "identity", alpha = 0.6, color = "black") +
  labs(x = "Log Antibody Level", y = "Count", title = "Distribution of Log Antibody Levels by Race/Ethnicity") +
  scale_fill_manual(values = c("blue", "orange", "green", "purple"), 
                    labels = c("1" = "White", "2" = "Asian", "3" = "Black", "4" = "Hispanic")) +
  theme_minimal() +
  theme(legend.title = element_blank())

race_antibody

#Average Log Antibody Levels by Smoking Status
smoking_antibody <- ggplot(table_data, aes(x = factor(smoking), y = log_antibody)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(x = "Smoking Status", y = "Average Log Antibody Level", 
       title = "Average Log Antibody Levels by Smoking Status") +
  scale_x_discrete(labels = c("0" = "Never Smoked", 
                               "1" = "Former Smoker", 
                               "2" = "Current Smoker")) +
  theme_minimal()

smoking_antibody

#Distribution of Log Antibody Levels by BMI 
bmi_antibody <- ggplot(table_data, aes_string(x = "bmi", y = "log_antibody")) +
  geom_point(color = "darkgreen", alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.5, color = "red", se = FALSE) +
  theme_bw() +
  labs(x = "BMI", y = "Log Antibody Levels")

bmi_antibody

#Average Log Antibody levels by diabetes status
diabetes_antibody <- ggplot(table_data, aes(x = factor(diabetes), y = log_antibody)) +
  geom_boxplot() +
  labs(x = "Diabetes", y = "Log Antibody Level") +
  theme_minimal()

diabetes_antibody

#Average Log Antibody levels by hypertension status
hypertension_antibody <- ggplot(table_data, aes(x = factor(hypertension), y = log_antibody)) +
  geom_boxplot() +
  labs(x = "Hypertension", y = "Log Antibody Level") +
  theme_minimal()

hypertension_antibody

#Distribution of Log Antibody Levels by SBP 
sbp_antibody <- ggplot(table_data, aes_string(x = "SBP", y = "log_antibody")) +
  geom_point(color = "darkgreen", alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.5, color = "red", se = FALSE) +
  theme_bw() +
  labs(x = "SBP", y = "Log Antibody Levels")

sbp_antibody 

#Distribution of Log Antibody Levels by LDL
LDL_antibody <- ggplot(table_data, aes_string(x = "LDL", y = "log_antibody")) +
  geom_point(color = "darkgreen", alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.5, color = "red", se = FALSE) +
  theme_bw() +
  labs(x = "LDL", y = "Log Antibody Levels")

LDL_antibody

#Distribution of Log Antibody Levels by Time Since Vaccination 
time_antibody <- ggplot(table_data, aes_string(x = "time", y = "log_antibody")) +
  geom_point(color = "darkgreen", alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.5, color = "red", se = FALSE) +
  theme_bw() +
  labs(x = "Time Since Vaccination", y = "Log Antibody Levels")

time_antibody
```

```{r Bivariate analysis}
#age & antibody (p <2e-16 for intercept and age)
b <- lm(log_antibody ~ age, data = table_data)
summary(b) 

#gender & antibody (t = 17.523, df = 4972.7, p-value < 2.2e-16)
t.test(log_antibody ~ gender, data = table_data)

#race & antibody (Pr(>F) = 0.367)
anova_race <- aov(log_antibody ~ race, data = table_data)
summary(anova_race)

#smoking & antibody (Pr(>F) = 1.95e-10)
anova_smoke <- aov(log_antibody ~ smoking, data = table_data)
summary(anova_smoke)

# bmi & antibody (p-value: < 2.2e-16)
bmi_linear <- lm(log_antibody ~ bmi, data = table_data)
summary(bmi_linear) 

#diabetes & antibody p-value = 0.6828
t.test(log_antibody ~ diabetes, data = table_data)

#hypertension & antibody p-value = 5.491e-05
t.test(log_antibody ~ hypertension, data = table_data) 

#sbp and antibody p-value: 1.451e-05
sbp_linear <- lm(log_antibody ~ SBP, data = table_data)
summary(sbp_linear)  

#LDL and antibody p-value: 0.01143
LDL_linear <- lm(log_antibody ~ LDL, data = table_data)
summary(LDL_linear)

#time and antibody p-value: 0.3278
time_linear <- lm(log_antibody ~ time, data = table_data)
summary(time_linear)
```


```{r the CART approach}
#setting seed for reproducibility
set.seed(1)

#applying the regression tree method to the data using a complexity parameter of 0.01
tree1 = rpart(formula = log_antibody~.,
              data = dat1,
              control = rpart.control(cp=0.01))

rpart.plot(tree1)

#applying cost complexity pruning to obtain a tree with the right size
printcp(tree1)
cpTable = tree1$cptable
plotcp(tree1)

#picking the cp that yields the minimum cross-validation error
minErr = which.min(cpTable[,4])
tree3 = rpart::prune(tree1, cp = cpTable[minErr,1])
rpart.plot(tree3)
plot(as.party(tree3))

#predictions on the test data set
head(predict(tree3, newdata=dat2))

#computing the RMSE on the test set
RMSE(predict(tree3, newdata=dat2),dat2$log_antibody)
```
The RMSE for the regression tree model is 0.5873775.

```{r Model Building TINA }
x <- model.matrix(log_antibody ~ ., dat1) [, -1]
y <- dat1$log_antibody

x2 <- model.matrix(log_antibody ~ ., dat2) [, -1]
y2 <- dat2$log_antibody

ctrl1 = trainControl(method = "cv", number = 10)

#GAM 
gam.fit <- train(x, y,
method = "gam",
trControl = ctrl1)

gam.fit$bestTune
gam.fit$finalModel
gam.m1 <- gam(log_antibody ~ time + age + gender + race + smoking + bmi + diabetes + hypertension + 
    SBP + LDL, data = dat1) 

gam.m2 <- gam(log_antibody ~ s(time) + age + gender + race + smoking + bmi + diabetes + hypertension + 
    SBP + LDL, data = dat1) 

anova(gam.m1, gam.m2, test = "F")

plot(gam.m2)
#Summary of the model
summary(gam.fit)

predy2_gam2 = predict(gam.fit, newdata=x2)
mean((y2 - predy2_gam2)^2)

#MARS
mars_grid <- expand.grid(degree = 1:3,
                         nprune = 2:24) 
set.seed(2)

mars.fit <- train(x, y,
                  method = "earth",
                  tuneGrid = mars_grid,
                  trControl = ctrl1)
print(mars.fit)

predy2_mars2 = predict(mars.fit, newdata=x2)
mean((y2 - predy2_mars2)^2)

#comparing models based on resampling results
resamp = resamples(list(gam = gam.fit,
                        mars = mars.fit))
bwplot(resamp, metric = "RMSE")
```
The RMSE for the MARS model is 0.2838458. The RMSE for the GAM model is 0.3233601. Resampling results provide additional evidence that the strongest of the two models compared above is the MARS model.

```{r}
ctrl1 = trainControl(method = "repeatedcv",
                     number = 10,
                     repeats = 5,
                     selectionFunction = "best")

#pcr using caret
set.seed(2)

pcr_fit = train(x,y,
                method = "pcr",
                tuneGrid = data.frame(ncomp=1:5),
                trControl = ctrl1,
                preProcess = c("center","scale"))

predy2_pcr2 = predict(pcr_fit, newdata=x2)
mean((y2 - predy2_pcr2)^2)

#pls using caret
pls_fit = train(x,y,
                method="pls",
                tunegrid = data.frame(ncomp=1:5),
                trControl = ctrl1,
                preProcess = c("center","scale"))

predy2_pls2 = predict(pls_fit, newdata=x2)
mean((y2 - predy2_pls2)^2)

#enet using caret
enet_fit = train(log_antibody ~.,
                 data = dat1,
                 method = "glmnet",
                 tuneGrid = expand.grid(alpha=seq(0,1, length = 21),
                                        lambda = exp(seq(6,0,length = 100))),
                 trControl = ctrl1)

#comparing models based on resampling results
resamp = resamples(list(pcr = pcr_fit,
                        pls = pls_fit,
                        enet = enet_fit))
bwplot(resamp, metric = "RMSE")
```

The RMSE for the PCR model is 0.3675697. The RMSE for the PLS model is 0.324535. Resampling results provide additional evidence that the strongest of the three models compared above is the PLS model. 

```{r}

ctrl1 = trainControl(method = "cv", number = 10)

#lasso using caret
set.seed(2)
lasso.fit = train(log_antibody ~.,
                  data = dat1,
                  method = "glmnet",
                  tuneGrid = expand.grid(alpha=1,
                                         lambda = exp(seq(6,0,length=100))),
                  trControl = ctrl1)

#lm using caret
lm.fit = train(log_antibody ~.,
               data = dat1,
               method = "lm",
               trControl = ctrl1)

#ridge using caret
ridge.fit = train(log_antibody ~.,
                  data = dat1,
                  method = "glmnet",
                  tuneGrid = expand.grid(alpha=1,
                                         lambda = exp(seq(6,0,length=100))),
                  trControl = ctrl1)

#comparing models based on resampling results
resamp = resamples(list(lasso = lasso.fit,
                        lm = lm.fit,
                        ridge = ridge.fit))
bwplot(resamp, metric = "RMSE")

```
Among the three models compared above, the lm model appears to be the 'best' based on resampling results (as it has the lowest RMSE).  

